Synthesizing Quotient Lenses

### 2018-03-15 papers/QuotientLenses/main.tex
Made edits to abstract and intro

\begin{document}
\title{Synthesizing Quotient Lenses}
\begin{abstract}
{\em Quotient lenses} are bidirectional transformations whose correctness
laws are ``loosened'' by specified equivalence relations, allowing
inessential details in concrete data formats to be suppressed. 
For example, a programmer could use a quotient lens to define 
a transformation that ignores the order of fields in XML data, so
that two XML files with the same fields but in different orders would be
considered the same, allowing a single, simple program to handle them both. 
%
Building on a recently published algorithm for synthesizing plain bijective
lenses from high-level specifications, we show how to synthesize bijective
{\em quotient} lenses in three steps.  First, we introduce {\em quotient regular
  expressions} (QREs), annotated regular expressions that conveniently mark
inessential aspects of string data formatsl; each QRE specifies,
simulteneously, a regular language and an equivalence relation on it.
Second, we introduce {\em QRE lenses}, i.e., lenses mapping between QREs.
Our key technical result is a proof that every QRE lens can be transformed into
one of the form $c ; \ell ; d^{-1}$, where $c$ and $d$ are {\em
canonizers}, $\ell$ is a bijective transformation, and $;$ is
composition\bcp{This seems not quite accurate.  It makes it look like
  canonizers are bidirectional things, but they aren't---at least, we
don't talk about them in that way (I suppose we could define the ``other
direction'' to be the identity, but we do not).  What we really mean are
left quotient and right quotient (and then we don't need the $^{-1}$).
Similarly, the meaning of 
``;'' here is not sequential composition of either lenses or QREs, which are
the only two meanings we give to this symbol in the paper.}.  
% Intuitively,
% these canonizers normalize data by 
% picking a canonical representative from the relevant equivalence
% relations. 
% BCP: Don't think this adds much, and it gets between the theorem and its use.
This theorem means we can push all normalization to the outer edges of a
composite quotient lens, leaving in the middle a pure bijective
transformation. Third, we leverage this normalization theorem to {\em
synthesize} QRE lenses from a pair of QREs and example input-output pairs,
reusing earlier work on synthesizing plain bijective lenses. We have
implemented QREs and QRE lens synthesis as an extension to the bidirectional programming
language Boomerang. We evaluate the effectiveness of our approach by 
synthesizing QRE lenses between various real-world data formats in the
Optician benchmark suite.

\end{abstract}

@@

Our second main contribution in this paper is to introduce {\em QRE lenses},
which are bijective quotient lenses that map between QREs via bijective lenses
that map between the respective canonical formats defined by the source and
target QREs. This idea is simple and natural, but begs the following question:
do we give up expressiveness if we restrict ourselves to this form? Our main
technical contribution to this question asserts that we do not---more
specifically, we prove a normal form theorem (Theorem~\ref{normal form})
that says that every QRE lens can be rewritten to be the composition of a
source canonizer, a bijective lens, and a target canonizer. This normalization
property will enable us to (1) synthesize QRE lenses by extending the synthesis
algorithm used by Optician, and (2) prove that if there is a QRE lens that
satisfies the input specification, then this extended algorithm will return
such a lens.

@@

Indeed this idea is what motivates our definition of {\em QRE lenses}.
Intuitively, our QRE lenses are bijective lenses with ``canonizers at the edges,''
just like Boomerang's quotient lenses. 
Figure~\ref{fig:attheedges} depicts the architecture of QRE lenses.
\begin{figure}[t]
\centering
\includegraphics[width=0.7\textwidth]{QREs}
\caption{QRE lens from source $P$ to target $Q$.  $P$ and $Q$ are QREs, each
  consisting of a ``whole'' set ($W(P)$ and $W(Q)$) and a ``kernel'' part
  ($K(P)$ and $K(Q)$) Between the two kernels is a plain bijective lens.
  The \kw{get} function of the whole lens takes an argument from $W(P)$,
  applies the canonizer for $P$ to obtain a canonical representative in
  $K(P)$, then applies the \kw{get} of the plain lens, yielding an element
  of $K(Q)$, which is a subset of $W(Q)$.  The \kw{put} function does the
  reverse, mapping from $W(Q)$ to $K(P)$ (hence $W(P)$.  This lens is in
  normal form, with canonizers (specified by QREs) at the outer edges and an
  ordinary bijective lens in the center. \BCP{Both the figure and the
    caption are new---what do you think??}}\saz{I think that this figure is a
  big improvement.  One tiny possible tweak: color the \kw{get} and \kw{put}
  keywords in the same color as in the text.}\bcp{Can do, if we keep the
  color in the text :-).  How are we going to decide about that?}\saz{Vote?}
\label{fig:attheedges}
\end{figure}
Every QRE lens $q$ has a type $P \Leftrightarrow Q$ where $P$ and $Q$
are QREs. In the {\em get} direction, a QRE lens $q: P \Leftrightarrow Q$ uses the
source QRE $P$ to compute a canonical representative for the data modulo the
equivalence relation defined by $P$ and then applies the \kw{get} function of a
bijective string lens $\ell$ to this representative. In the {\em put} direction, $q$
operates similarly, but using the QRE $Q$ and the $\lput$ function of
$\ell$.

@@

The \kw{Formless}\saz{Is this ``Formless'' ``Formlens'' or ``Formlet''? The code
  below says ``Formlens'' but the earlier parts of the paragraph say ``Formlet''}type is given by
\begin{center}
\begin{lstlisting}
type Formlens a = Maybe a $\to$ [Int] $\to$ (Html,Env $\to$ Maybe a,[Int])
\end{lstlisting}
\end{center}
with the differences between this definition and the standard definition for
formlets being an extra \codefont{Maybe\; a} parameter that provides an
optional initial value of the form, the namesource being changed to a list of
integers, and the {\em collect} function being allowed to return an optional
value.

@@

\BCP{Wrote the rest---comments?}\saz{I swapped the last two paragraphs to end on
  a more exciting note.}
Looking ahead, we are experimenting with larger examples of
quotiented bijective lenses.  So far, the three ``specialized'' QRE
combinators (\kw{squash}, \kw{perm}, and \kw{collapse}) have proved
sufficient, but we expect eventually to encounter examples that will lead us
to consider adding further combinators to supplement these.

Along a different dimension, we would like to investigate whether the techniques
proposed here can be applied to synthesizing quotiented variants of richer
classes of lenses, such as ``classic'' asymmetric lenses~\cite{FosterLenses} or
symmetric lenses~\cite{symmetriclenses}.  The main technical issues are
developing synthesis procedures for the unquotientied variants of a given
class---we are currently attacking the case of symmetric lenses---and showing
that quotiented lenses in this class can be normalized along the lines of
Theorem~\ref{normal form}.

>>>>

\begin{document}
\title{Synthesizing Quotient Lenses}
\begin{abstract}
{\em Quotient lenses} are bidirectional transformations whose correctness
laws are ``loosened'' by specified equivalence relations, allowing
inessential details in concrete data formats to be suppressed. 
For example, a programmer could use a quotient lens to define 
a transformation that ignores the order of fields in XML data, so
that two XML files with the same fields but in different orders would be
considered the same, allowing a single, simple program to handle them both. 
%
Building on a recently published algorithm for synthesizing plain bijective
lenses from high-level specifications, we show how to synthesize bijective
quotient lenses in three steps. First, we introduce {\em quotient regular
  expressions} (QREs), annotated regular expressions that conveniently mark
inessential aspects of string data formats; each QRE specifies,
simulteneously, a regular language and an equivalence relation on it.
Second, we introduce {\em QRE lenses}, i.e., lenses mapping between QREs.
Our key technical result is a proof that every QRE lens can be transformed
into a functionally equivalent lens that canonizes source and target data only
once at the ``edges'' then uses a bijective lens to map between the respective
canonical elements; no internal canonization occurs in a lens in this normal
form. Third, we leverage this normalization theorem to {\em synthesize} QRE
lenses from a pair of QREs and example input-output pairs, reusing earlier work
on synthesizing plain bijective lenses. We have implemented QREs and QRE lens
synthesis as an extension to the bidirectional programming language Boomerang.
We evaluate the effectiveness of our approach by synthesizing QRE lenses
between various real-world data formats in the Optician benchmark suite.
\end{abstract}

@@

% Our end goal is to synthesize bijective quotient lenses that map between QREs
% via a bijective lens between the respective canonical formats of the QREs. This
% idea is simple and natural, but begs the following question: do we give up
% expressiveness if we restrict ourselves to this form? We will show that this
% is not the case by doing the following. First, we will introduce {\em QRE
% lenses}, which are bijective quotient lenses formed by ``lifting'' bijective
% lenses to a bijective quotient lens type, then freely quotienting these lenses
% by QREs or applying familiar regular and composition operators to them. Then,
% we will prove a normal form theorem (Theorem~\ref{normal form})
% that says that every QRE lens can be transformed into a functionally equivalent
% lens that canonizes source and target data only once at the ``edges'' then uses
% a bijective lens to map between the respective canonical elements; no internal
% canonization occurs in a lens in this normal form. This normalization
% property will enable us to (1) synthesize QRE lenses by extending the synthesis
% algorithm used by Optician, and (2) prove that if there is a QRE lens that
% satisfies the input specification, then this extended algorithm will return
% such a lens.

Our second main contribution in this paper is to introduce {\em QRE lenses},
with the end goal of synthesizing QRE lenses that map between QREs via a
synthesized bijective lens between the respective canonical formats. This idea
is simple and natural but begs the following question: do we give up
expressiveness if we restrict ourselves to this form? Our main technical
contribution to this question asserts that we do not---more specifically, we
prove a normal form theorem (Theorem~\ref{normal form}) that says that every
QRE lens formed by freely composing QRE lenses, applying regular
operators to them, and quotenting them with QREs can be rewritten to be the
composition of a source canonizer, a bijective lens, and a target canonizer.
This normalization property will enable us to (1) synthesize QRE lenses by
extending the synthesis algorithm used by Optician, and (2) prove that if there
is a QRE lens that satisfies the input specification, then this extended
algorithm will return such a lens.

@@

Indeed this idea is what motivates our definition of {\em QRE lenses}.
Intuitively, our QRE lenses are bijective lenses with ``canonizers at the
edges''.
Figure~\ref{fig:attheedges} depicts the architecture of QRE lenses.
\begin{figure}[t]
\centering
\includegraphics[width=0.7\textwidth]{QREs}
\caption{QRE lens from source $P$ to target $Q$.  $P$ and $Q$ are QREs, each
  consisting of a ``whole'' set ($W(P)$ and $W(Q)$) and a ``kernel'' part
  ($K(P)$ and $K(Q)$) Between the two kernels is a plain bijective lens.
  The \kw{get} function of the whole lens takes an argument from $W(P)$,
  applies the canonizer for $P$ to obtain a canonical representative in
  $K(P)$, then applies the \kw{get} of the plain lens, yielding an element
  of $K(Q)$, which is a subset of $W(Q)$.  The \kw{put} function does the
  reverse, mapping from $W(Q)$ to $K(P)$ (hence $W(P)$.  This lens is in
  normal form, with canonizers (specified by QREs) at the outer edges and an
  ordinary bijective lens in the center. \BCP{Both the figure and the
    caption are new---what do you think??}}\saz{I think that this figure is a
  big improvement.  One tiny possible tweak: color the \kw{get} and \kw{put}
  keywords in the same color as in the text.}\bcp{Can do, if we keep the
  color in the text :-).  How are we going to decide about that?}\saz{Vote?}
\label{fig:attheedges}
\end{figure}
Every QRE lens $q$ has a type $P \Leftrightarrow Q$ where $P$ and $Q$
are QREs. In the {\em get} direction, a QRE lens $q: P \Leftrightarrow Q$ uses
the source QRE $P$ to compute a canonical representative for the data modulo
the equivalence relation defined by $P$ and then applies the \kw{get} function
of a bijective string lens $\ell$ to this representative. In the {\em put}
direction, $q$ operates similarly, but using the QRE $Q$ and the $\lput$
function of $\ell$.

@@

The Formlens type is given by
\begin{center}
\begin{lstlisting}
type Formlens a = Maybe a $\to$ [Int] $\to$ (Html,Env $\to$ Maybe a,[Int])
\end{lstlisting}
\end{center}
with the differences between this definition and the standard definition for
formlets being an extra \codefont{Maybe\; a} parameter that provides an
optional initial value of the form, the namesource being changed to a list of
integers, and the {\em collect} function being allowed to return an optional
value.

@@

Looking ahead, we are experimenting with larger examples of
quotiented bijective lenses.  So far, the three ``specialized'' QRE
combinators (\kw{squash}, \kw{perm}, and \kw{collapse}) have proved
sufficient, but we expect eventually to encounter examples that will lead us
to consider adding further combinators to supplement these.

Along a different dimension, we would like to investigate whether the techniques
proposed here can be applied to synthesizing quotiented variants of richer
classes of lenses, such as ``classic'' asymmetric lenses~\cite{Lenses} or
symmetric lenses~\cite{hofmann2011symmetric}.  The main technical issues are
developing synthesis procedures for the unquotientied variants of a given
class---we are currently attacking the case of symmetric lenses---and showing
that quotiented lenses in this class can be normalized along the lines of
Theorem~\ref{normal form}.
